RL-Musician: A Configurable Open-Source Tool for Species Counterpoint Composition

### 2020-04-14 docs/paper/paper.tex
docs: Update conclusion of a paper draft

\date{April 4, 2020}

@@

Currently, there are no perfect tools for algorithmic composition. Plenty of various approaches for composing music automatically exist, but none of them produce well-structured and novel pieces that cannot be distinguished from works of a talented human composer. Thus, there is an open research problem. Several recent breakthroughs in it are accomplished with machine learning and both supervised learning \cite{johnson2017generating, payne2019musenet} and reinforcement learning \cite{yi2007automatic, smith2012reinforcement} are applicable. Moreover, some researchers combine them \cite{jaques2016generating, kotecha2018bach, kumar2019polyphonic}. Meanwhile, fitness function optimization also yields prominent results\cite{herremans2012first, herremans2013fifth}.

@@

These examples can be divided into two groups. The first one is formed by approaches where given fitness function is optimized.  The second group is formed by composition of music with reinforcement learning in its narrow sense (i.e., methods based on value function estimation). This group is something in between the first group and supervised learning approaches. On the one hand, reward function is a domain-specific name for fitness function. On the other hand, data generated during past episodes are used for training supervised models.

@@

Consider a tree search problem. To be more detailed, suppose that there is a tree such that every its leaf is mapped to a reward and the goal is to find the leaf with the maximum reward. However, leaves and corresponding to them rewards are not known initially. Information about a leaf becomes available only after a path from the root of the tree to this leaf is constructed. %In other words, the goal is to find optimum sequence of nodes starting at the root and ending at a leaf.

@@

\begin{itemize}
	\item \textbf{Absence of looped pitches.} A stable pitch may be rearticulated, but not more than once in a row. Every excessive occurrence of the same pitch within a line adds constant negative reward.
    \item \textbf{Absence of looped fragments.} Define a fragment as one or more consecutive measures (at a piece level, not at a melodic line level). For every fragment followed by exactly the same fragment, a constant negative reward is added.
	\item \textbf{Motion.} For every pair of lines every motion between two successive measures adds a reward that depends on a type of motion. Since contrary motion makes lines more distinguishable, reward for it is positive. Reward for oblique motion is zero, reward for similar motion is slightly negative and reward for parallel motion is negative, because parallel motion makes lines sounding like copies of each other.
	\item \textbf{Correlation between lines.} Consider each line as a sequence of real numbers equal to pitch positions within a range of available pitches. Then average over all pairs of lines correlation between two lines can quantify independence of lines. Actually, reward is rescaled to be from 0 to 1 -- it is half of difference of 1 and the above average correlation.
    \item \textbf{Explicity of climax.} Music theory suggests that good melodies must have goal-oriented motion and that such goal-orientedness may be achieved by having exactly one climax point. For each line, constant reward is decreased for each duplication of climax point and for not so high climax point.
    \item \textbf{Number of skips.} Positive reward is associated with every line such that number of skips in it lies within a specified range. The motivation behind this is that lines with almost no skips are not interesting, but lines with too many skips are not coherent.
    \item \textbf{Absence of downward skips}. Counterpoint theory prescribes to have downward step motion. To encourage its usage, there are negative rewards for every downward skip and the large a skip is, the higher a penalty its.
\end{itemize}

@@

The results look like scores for climax explicity and absence of looped pitches, absence of looped fragments, and absence of downward skips impose hard constraints. If so, the goal is to maximize sum of scores for lines independence subject to counterpoint rules and above new constraints.

Table \ref{table:piece} contains found by Monte Carlo Beam Search piece for which the scores are reported.

@@

This paper introduces a new approach for music composition without usage of existing pieces.

The list of suggested further improvements is as follows:
\begin{itemize}
	\item Use 
\end{itemize}

>>>>

\date{April 14, 2020}

@@

Currently, there are no perfect tools for algorithmic composition. Plenty of various approaches to composing music automatically exist, but none of them produce well-structured and novel pieces that cannot be distinguished from works of a talented human composer. Thus, there is an open research problem. Several recent breakthroughs in it are accomplished with machine learning and both supervised learning \cite{johnson2017generating, payne2019musenet} and reinforcement learning \cite{yi2007automatic, smith2012reinforcement} are applicable. Moreover, some researchers combine them \cite{jaques2016generating, kotecha2018bach, kumar2019polyphonic}. Meanwhile, fitness function optimization also yields prominent results\cite{herremans2012first, herremans2013fifth}.

@@

These examples can be divided into two groups. The first one is formed by approaches where an explicitly given fitness function is optimized. The second group is formed by composition of music with reinforcement learning in its narrow sense (i.e., methods based on value function estimation). This group is something in between the first group and supervised learning approaches. On the one hand, reward function is a domain-specific name for fitness function. On the other hand, data generated during past episodes are used for training supervised models.

@@

Consider a tree search problem. To be more detailed, suppose that there is a tree such that every its leaf is mapped to a reward and the goal is to find the leaf with the maximum reward. However, leaves and corresponding to them rewards are not known initially. Information about a leaf becomes available only after a path from the root of the tree to this leaf is constructed.

@@

\begin{itemize}
	\item \textbf{Absence of looped pitches.} A stable pitch may be rearticulated, but not more than once in a row. Every excessive occurrence of the same pitch within a line adds constant negative reward.
    \item \textbf{Absence of looped fragments.} Define a fragment as one or more consecutive measures (at a piece level, not at a melodic line level). For every fragment followed by exactly the same fragment, a constant negative reward is added.
	\item \textbf{Motion.} For every pair of lines every motion between two successive measures adds a reward that depends on a type of motion. Since contrary motion makes lines more distinguishable, reward for it is positive. Reward for oblique motion is zero, reward for similar motion is slightly negative and reward for parallel motion is negative, because parallel motion makes lines sounding like copies of each other.
	\item \textbf{Correlation between lines.} Consider each line as a sequence of real numbers equal to pitch positions within a range of available pitches. Then average over all pairs of lines correlation between two lines can quantify independence of lines. Actually, reward is rescaled to be from 0 to 1 -- it is half of difference of 1 and the above average correlation.
    \item \textbf{Explicity of climax.} Music theory suggests that good melodies must have goal-oriented motion and that such goal-orientedness may be achieved by having exactly one climax point. For each line, constant reward is decreased for each duplication of climax point and for not so high climax point.
    \item \textbf{Number of skips.} Positive reward is associated with every line such that number of skips in it lies within a specified range. The motivation behind this is that lines with almost no skips are not interesting, but lines with too many skips are not coherent.
    \item \textbf{Absence of downward skips}. Counterpoint theory prescribes to have downward step motion. To encourage its usage, there are negative rewards for every downward skip and the larger a skip is, the higher a penalty its.
\end{itemize}

@@

The results look like scores for climax explicity, absence of looped pitches, absence of looped fragments, and absence of downward skips impose hard constraints. If so, the goal is to maximize sum of scores for lines independence subject to counterpoint rules and above new constraints.

Table \ref{table:piece} demonstrates found by Monte Carlo Beam Search piece for which the scores are reported.

@@

This paper introduces a new approach to composition of music without usage of existing pieces. For a problem of first species counterpoint writing, the approach finds pieces that maximize weighted sum of associated with particular musical properties scores subject to boundary constraints. These boundary constraints (the first measure and the last measure) are user-defined and weights for the sum of scores are user-defined too.

Results are good from formal metrics point of view -- the method outperforms random search and manual trials. However, created pieces are too simple to be enjoyable music. This is an inevitable consequence of too strict limitations of first species counterpoint. To eliminate them, higher species can be chosen or frameworks other than species counterpoint can be explored.

If the problem becomes computationally complex after introducing more musical freedom, concepts from reinforcement learning such as value function can be involved. State-free search can be replaced with a search where decision are made based on observations and estimated values of states and actions.
